---
title: "Covid91 vaccine study Final2024"
author: "Quintin Burnett"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, covid19 = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
Male = filter(FinalData, sex =="M")
Female = filter(FinalData, sex =="F")
Gay = filter(FinalData, LGBTQ =="gay")
DrugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.


## Overall approach

1. **Descriptive Analysis**: Summarize data visually then numerically.
2. **Inferential Analysis**: Conduct hypothesis testing:
   - \(H_0\): The vaccine has no impact on infection rates.
   - \(H_A\): The vaccine reduces infection rates.
3. **Effectiveness Calculation**:
   \[
   \text{Vaccine Effectiveness (VE)} = 1 - \frac{\text{Risk}_{\text{vaccine}}}{\text{Risk}_{\text{placebo}}}
   \]
4. **Effectivness Confidence Interval**: low to hi
   
## Males

### Summary of Males
```{r}
male_summary <- Male %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
male_summary <- male_summary %>%
  mutate(total = rowSums(select(., -treatment)))
male_column_totals <- male_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
male_summary <- bind_rows(male_summary, male_column_totals)
male_summary
```

This is the given information so far for the males. There are 18392 males. 355 have covid while 18037 are OK. This isn't enough to show a trend so now let us add a barplot for a visual representation.

### Visual Results

```{r}
# Bar plot

# Summarize the data for Males by infection status and treatment
male_summary_data <- Male %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
male_plot_count <- ggplot(male_summary_data, aes(x = infected, y = count, fill = treatment))
male_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Males Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
male_plot_percent <- ggplot(male_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
male_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Males Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The Barplots above shows the number of males infected with covid. Visually males who take the vaccine have less covid than the placebo but the results are similarly close to the placebo on both graphs. Visually they are still different but we need to conduct a statistical analysis to confirm if the difference is significant.

### Numerical Results
```{r}
table1 <- xtabs(~infected + treatment, data=Male)
rowPerc(table1)
colPerc(table1)
```

The numerical results show that in the infected males, **73.8%** had taken the placebo, while **26.2%** had taken the vaccine. For those who are OK, **50.01%** had taken the placebo, and **49.99%** had taken the vaccine. Overall, **2.82%** of males on the placebo have covid compared to **1.02%** on the vaccine, indicating that the vaccine reduces infection rates. These findings suggest the vaccine is effective in lowering covid with the males. With this we can coorelate this to the barplots proving the effectiveness.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
Malefish <- fisher.test(table1)
Malefish
```

The key result to focus on is Fisher's Exact Test, where the p-value is extremely small (\(p = 2.2 \times 10^{-16}\)) or **0.00000000000000022**. This means the observed differences in infection rates between the vaccine and placebo groups are very unlikely to occur by chance. The null hypothesis, which assumes infection rates are independent of treatment, is rejected. The odds ratio of about **2.82** shows that males who received the vaccine were nearly **2.8** times more likely to get covid compared to those who received the placebo. These results indicate the vaccine is **ineffective** at preventing covid in males.

```{r}
# Calculate effectiveness
male_effect <- Malefish$p.value / (Malefish$p.value + 1)
effectiveness_male <- (1 - 1 / Malefish$estimate) * 100
print(effectiveness_male)

# Calculate effectiveness confidence interval
male_low <- (1 - 1 / Malefish$conf.int[1]) * 100
male_hi <- (1 - 1 / Malefish$conf.int[2]) * 100
options(digits = 2)

#print 
male_low
male_hi
```

### `r effectiveness_male`% Effectiveness with Confidence Interval: `r male_low`%  to `r male_hi`%

Above is the effectiveness of the covid vaccine for males. If the vaccine's effectiveness is above **50%**, it indicates that the vaccine is effective. Based on the calculated effectiveness of **64.48%** and the confidence interval ranging from **54.75%** to **72.32%**, we can conclude that the vaccine is **effective** in reducing infection rates among males.

The null hypothesis (\(H_0\)) assumed that the vaccine has no impact on infection rates, while the alternative hypothesis (\(H_A\)) posited that the vaccine reduces infection rates. Given the very small p-value from Fisher’s Exact Test and the calculated odds ratio of **2.82**, we reject the null hypothesis and accept the alternative, confirming that the vaccine does reduce infection rates in males.

The odds ratio indicates that males who received the vaccine were nearly **2.8** times more likely to avoid contracting covid compared to those who received the placebo. The confidence interval further supports this, showing a **95%** range of effectiveness between **54.75%**% and **72.32%**, reinforcing the vaccine's significant protective effect.

## Females

### Summary of Females
```{r}
female_summary <- Female %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
female_summary <- female_summary %>%
  mutate(total = rowSums(select(., -treatment)))
female_column_totals <- female_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
female_summary <- bind_rows(female_summary, female_column_totals)
female_summary
```

This is the given information so far for the females. There are 18328 females. 365 have covid while 17963 are OK. This isn't enough to show a trend so now let us add a barplot for a visual representation.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Females by infection status and treatment
female_summary_data <- Female %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
female_plot_count <- ggplot(female_summary_data, aes(x = infected, y = count, fill = treatment))
female_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Females Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
female_plot_percent <- ggplot(female_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
female_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Females Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The Barplots above shows the number of females infected with covid. Visually females who take the vaccine have less covid than the placebo but the results are similarly close to the placebo on both graphs. Visually they are still different but we need to conduct a statistical analysis to confirm if the difference is significant.

### Numerical Results
```{r}
table2 <- xtabs(~infected + treatment, data=Female)
rowPerc(table2)
colPerc(table2)
```

The numerical results show that in the infected females, **65%** had taken the placebo, while **35%** had taken the vaccine. For those who are OK, **50%** had taken the placebo, and **50%** had taken the vaccine. Overall, **2.6%** of females on the placebo have covid compared to **1.4%** on the vaccine, indicating that the vaccine reduces infection rates. These findings suggest the vaccine is effective in lowering covid rates in females. This aligns with the barplots showing the vaccine's effectiveness.

### Inferential Results
```{r}
chisq.test(table2)
chisqtestGC(table2)
Femalefish <- fisher.test(table2)
Femalefish
```

The key result to focus on is Fisher's Exact Test, where the p-value is extremely small (\(p = 1 \times 10^{-8}\)) or **0.00000001**, indicating that the observed differences in infection rates between the vaccine and placebo groups are very unlikely to have occurred by chance. The null hypothesis, which assumes infection rates are independent of treatment, is rejected. The odds ratio of about **1.9** shows that females who received the vaccine were nearly **1.9** times more likely to get covid compared to those who received the placebo. These results suggest that the vaccine is **ineffective** at preventing covid in females.

```{r}
# Calculate effectiveness
female_effect <- Femalefish$p.value / (Femalefish$p.value + 1)
effectiveness_female <- (1 - 1 / Femalefish$estimate) * 100
print(effectiveness_female)

# Calculate effectiveness confidence interval
female_low <- (1 - 1 / Femalefish$conf.int[1]) * 100
female_hi <- (1 - 1 / Femalefish$conf.int[2]) * 100
options(digits = 2)

#print 
female_low
female_hi
```

### `r effectiveness_female`% Effectiveness with Confidence Interval: `r female_low`%  to `r female_hi`%

Above is the effectiveness of the covid vaccine for females. If the vaccine's effectiveness is above **50%**, it indicates that the vaccine is effective. Based on the calculated effectiveness of **46.13%** and the confidence interval ranging from **32.79%** to **56.98%**, we can conclude that the vaccine is **ineffective** in reducing infection rates among females.

The null hypothesis (\(H_0\)) assumed that the vaccine has no impact on infection rates, while the alternative hypothesis (\(H_A\)) posited that the vaccine reduces infection rates. Given the very small p-value from Fisher’s Exact Test and the calculated odds ratio of **1.9**, we reject the null hypothesis and accept the alternative, confirming that the vaccine does reduce infection rates in females.

The odds ratio indicates that females who received the vaccine were nearly **1.9** times more likely to get covid compared to those who received the placebo. The confidence interval further supports this, showing a **95%** range of effectiveness between **32.79%** and **56.98%**, reinforcing that the vaccine's effect in females is limited.


## LGBTQ -> Gay

### Summary of Gay
```{r}
gay_summary <- Gay %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
gay_summary <- gay_summary %>%
  mutate(total = rowSums(select(., -treatment)))
gay_column_totals <- gay_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
gay_summary <- bind_rows(gay_summary, gay_column_totals)
gay_summary
```

This is the given information so far for the gays. There are 3060 gays. 60 have covid while 3000 are OK. This isn't enough to show a trend so now let us add a barplot for a visual representation.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Gay by infection status and treatment
gay_summary_data <- Gay %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
gay_plot_count <- ggplot(gay_summary_data, aes(x = infected, y = count, fill = treatment))
gay_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
gay_plot_percent <- ggplot(gay_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
gay_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The Barplots above shows the number of gays infected with covid. Visually gays who take the vaccine have less covid than the placebo. Visually they are very different but we need to conduct a statistical analysis to confirm if the difference is significant.

### Numerical Results
```{r}
table3 <- xtabs(~infected + treatment, data=Gay)
rowPerc(table3)
colPerc(table3)
```

The numerical results show that in the infected gays, **80%** had taken the placebo, while **20%** had taken the vaccine. For those who are OK, **48%** had taken the placebo, and **52%** had taken the vaccine. Overall, **3.2%** of gay individuals on the placebo have covid compared to **0.77%** on the vaccine, indicating that the vaccine reduces infection rates. These findings suggest the vaccine is effective in lowering covid rates in gays. This aligns with the barplots showing the vaccine's effectiveness.

### Inferential Results
```{r}
chisq.test(table3)
chisqtestGC(table3)
Gayfish <- fisher.test(table3)
Gayfish
```

The key result to focus on is Fisher's Exact Test, where the p-value is extremely small (\(p = 7 \times 10^{-7}\)) or **0.0000007**, indicating that the observed differences in infection rates between the vaccine and placebo groups are very unlikely to have occurred by chance. The null hypothesis, which assumes infection rates are independent of treatment, is rejected. The odds ratio of about **4.3** shows that gay individuals who received the vaccine were nearly **4.3** times more likely to get covid compared to those who received the placebo. These results suggest that the vaccine is **ineffective** at preventing covid in gays.

```{r}
# Calculate effectiveness
gay_effect <- Gayfish$p.value / (Gayfish$p.value + 1)
effectiveness_gay <- (1 - 1 / Gayfish$estimate) * 100
print(effectiveness_gay)

# Calculate effectiveness confidence interval
gay_low <- (1 - 1 / Gayfish$conf.int[1]) * 100
gay_hi <- (1 - 1 / Gayfish$conf.int[2]) * 100
options(digits = 2)

#print 
gay_low
gay_hi
```

### `r effectiveness_gay`% Effectiveness with Confidence Interval: `r gay_low`%  to `r gay_hi`%

Above is the effectiveness of the covid vaccine for gays. If the vaccine's effectiveness is above **50%**, it indicates that the vaccine is effective. Based on the calculated effectiveness of **76.6%** and the confidence interval ranging from **55.06%** to **88.74%**, we can conclude that the vaccine is **effective** in reducing infection rates among gay individuals.

The null hypothesis (\(H_0\)) assumed that the vaccine has no impact on infection rates, while the alternative hypothesis (\(H_A\)) posited that the vaccine reduces infection rates. Given the very small p-value from Fisher’s Exact Test and the calculated odds ratio of **4.3**, we reject the null hypothesis and accept the alternative, confirming that the vaccine does reduce infection rates in gay individuals.

The odds ratio indicates that gay individuals who received the vaccine were nearly **4.3** times more likely to get covid compared to those who received the placebo. The confidence interval further supports this, showing a **95%** range of effectiveness between **55.06%** and **88.74%**, reinforcing the vaccine's significant protective effect for gays.

## Druggies -> DrugUser

### Summary of DrugUser
```{r}
druguser_summary <- DrugUser %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)

# Add total column
druguser_summary <- druguser_summary %>%
  mutate(total = rowSums(select(., -treatment)))

# Add total row for each treatment
druguser_column_totals <- druguser_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")

# Combine the totals with the main data
druguser_summary <- bind_rows(druguser_summary, druguser_column_totals)

druguser_summary
```

This is the given information so far for the drug users. There are 3060 drug users. 60 have covid while 3000 are OK. This isn't enough to show a trend so now let us add a barplot for a visual representation.

### Visual Results
```{r}
# Summarize the data for DrugUser by infection status and treatment
druguser_summary_data <- DrugUser %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
druguser_plot_count <- ggplot(druguser_summary_data, aes(x = infected, y = count, fill = treatment))
druguser_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
druguser_plot_percent <- ggplot(druguser_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
druguser_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The barplots above show the number of drug users infected with covid. Visually, drug users who take the vaccine have more covid compared to those on the placebo. While the difference appears significant visually, we need to conduct a statistical analysis to confirm if the difference is statistically significant.

### Numerical Results
```{r}
table4 <- xtabs(~infected + treatment, data=DrugUser)
rowPerc(table4)
colPerc(table4)
```

The numerical results show that among infected drug users, **8.3%** had taken the placebo, while **91.7%** had taken the vaccine. For those who are OK, **51.0%** had taken the placebo, and **49.0%** had taken the vaccine. Overall, **0.33%** of drug users on the placebo contracted covid compared to **3.6%** on the vaccine, indicating that the vaccine was associated with a higher infection rate in this group. These findings suggest that the vaccine may not be effective in lowering covid rates among drug users. This aligns with the barplots, which visually depict the higher infection rates among vaccinated drug users.

### Inferential Results
```{r}
chisq.test(table4)
chisqtestGC(table4)
DrugUserfish <- fisher.test(table4)
DrugUserfish
```

The key result to focus on is Fisher's Exact Test, where the p-value is extremely small (\(p = 3 \times 10^{-12}\)) or **0.000000000003**, indicating that the observed differences in infection rates between the vaccine and placebo groups are very unlikely to have occurred by chance. The null hypothesis, which assumes infection rates are independent of treatment, is rejected.

The odds ratio is approximately **0.087**, showing that drug users who received the vaccine were **11.5** times less likely to get covid compared to those who received the placebo. These results suggest that the vaccine is **highly effective** at preventing covid in drug users. This aligns with the numerical results showing reduced infection rates among vaccinated individuals in this population.

```{r}
# Calculate effectiveness
DrugUserfish <- fisher.test(table4)
effectiveness_drug <- (1 - DrugUserfish$estimate) * 100
print(effectiveness_drug)

# Calculate effectiveness confidence level
drug_low <- (1 - DrugUserfish$conf.int[2]) * 100  
drug_hi <- (1 - DrugUserfish$conf.int[1]) * 100  
options(digits = 2)

#print
drug_low
drug_hi
```

### `r effectiveness_drug`% Effectiveness with Confidence Interval: `r drug_low`%  to `r drug_hi`%

Above is the effectiveness of the covid vaccine for drug users. If the vaccine's effectiveness is above **50%**, it indicates that the vaccine is effective. Based on the calculated effectiveness of **91.3%** and the confidence interval ranging from **78.0%** to **97.3%**, we can conclude that the vaccine is **highly effective** in reducing infection rates among drug users.

The null hypothesis (\(H_0\)) assumed that the vaccine has no impact on infection rates, while the alternative hypothesis (\(H_A\)) posited that the vaccine reduces infection rates. Given the very small p-value from Fisher’s Exact Test and the calculated odds ratio of **0.087**, we reject the null hypothesis and accept the alternative, confirming that the vaccine does reduce infection rates in drug users.

The odds ratio of **0.087** indicates that drug users who received the vaccine were approximately **11.5** times less likely to get covid compared to those who received the placebo. Since the odds ratio is less than **1**, this means the vaccine has the **maximum possible effectiveness**. The confidence interval further supports this, showing a **95%** range of effectiveness between **78.0%** and **97.3%**, reinforcing the vaccine's significant protective effect for drug users.

# Overall Results and Conclusions

## Conclusion for Males

The covid vaccine is effective for males, providing **64.48% effectiveness** with a confidence interval ranging from **54.75% to 72.32%**. Statistical tests strongly support this conclusion:

- Fisher's Exact Test showed a highly significant p-value (\(p = 2.2 \times 10^{-16}\)), rejecting the null hypothesis of no vaccine impact.
- The odds ratio of **2.82** indicates that males who received the vaccine were nearly **2.8 times less likely** to get infected compared to those on the placebo.

Overall, the vaccine significantly reduces the likelihood of infection for males, supporting its effectiveness in this group.

## Conclusion for Females

## Conclusion for LGBTQ

## Conclusion for Druggies
