---
title: "Covid91 vaccine study Final2024"
author: "Quintin Burnett"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, covid19 = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
Male = filter(FinalData, sex =="M")
Female = filter(FinalData, sex =="F")
Gay = filter(FinalData, LGBTQ =="gay")
DrugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.


## Overall approach

1. **Descriptive Analysis**: Summarize data visually then numerically.
2. **Inferential Analysis**: Conduct hypothesis testing:
   - \(H_0\): The vaccine has no impact on infection rates.
   - \(H_A\): The vaccine reduces infection rates.
3. **Effectiveness Calculation**:
   \[
   \text{Vaccine Effectiveness (VE)} = 1 - \frac{\text{Risk}_{\text{vaccine}}}{\text{Risk}_{\text{placebo}}}
   \]
4. **Effectivness Confidence Interval**: low to hi
   
## Males

### Summary of Males
```{r}
male_summary <- Male %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
male_summary <- male_summary %>%
  mutate(total = rowSums(select(., -treatment)))
male_column_totals <- male_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
male_summary <- bind_rows(male_summary, male_column_totals)
male_summary
```

This is the given information so far for the males. There are 18392 males. 355 have covid while 18037 are OK. This isn't enough to show a trend so now let us add a barplot for a visual representation.

### Visual Results

```{r}
# Bar plot

# Summarize the data for Males by infection status and treatment
male_summary_data <- Male %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
male_plot_count <- ggplot(male_summary_data, aes(x = infected, y = count, fill = treatment))
male_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Males Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
male_plot_percent <- ggplot(male_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
male_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Males Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The Barplots above shows the number of males infected with covid. Visually males who take the vaccine have less covid than the placebo but the results are similarly close to the placebo on both graphs. Visually they are still different but we need to conduct a statistical analysis to confirm if the difference significant.

### Numerical Results



### Inferential Results



### Effectivness Results: low to high



## Females

### Summary of Females



### Visual Results



### Numerical Results



### Inferential Results



### Effectivness Results: low to high



## LGBTQ

### Summary of LGBTQ



### Visual Results



### Numerical Results



### Inferential Results



### Effectivness Results: low to high



## Druggies

### Summary of Druggies



### Visual Results



### Numerical Results



### Inferential Results



### Effectivness Results: low to high



# Overall Results and Conclusions